---
title: "Analysing US Airport Temperature Trends"
output: 
  html_notebook: 
    number_sections: no
    toc: yes
---


# Introduction

This is an analysis looking at long-term trends in temperature recorded at US airport stations. I orginally did this about a year ago, but am re-visiting it in more detail and with improved R and data science skills.




# Data

- In original analysis, I had downloaded all the files to my computer and was loading/combining them from there.
- I have now downloaded and cleaned all the data, and put into a sqlite database that I will query from instead of reading in the csv files. 
- See <https://github.com/andypicke/wunderground_data>







# Methods

- My plan is to fit a linear regression to long-term temperature records at each airport station.
- I will try doing this a variety of ways: Using all (daily) data, and using monthly/yearly averaged data.
- I will display the results (the slope of the linear regression) on a leaflet map, with the size and color of points mapped to the magnitude of the trend.



## Figure out how to query the data for 1 station from the database
- <http://www.datacarpentry.org/R-ecology-lesson/05-r-and-databases.html>
- Note The src_sqlite() command does not load the data into the R session ; use *collect* 
```{r}

library(ggplot2)
suppressPackageStartupMessages( library(dplyr) )

# open connection to sql database
db_path <- '/Users/Andy/Projects/wunderground_data/wunderground_daily.db'
wea_db <- src_sqlite(db_path)
wea_db
```

## Example Querying w/ SQL
```{r}
tbl(wea_db, sql("SELECT * FROM wea_clean WHERE mean_temp NOT NULL LIMIT 10"))
```



## Example querying w/ dplyr
```{r}
wea_df <- tbl(wea_db,'wea_clean') %>% 
        filter(st_code=='KDEN' & between(year,2008,2010) & !is.na(yday)) %>% 
        select(date,year,yday,mean_temp) %>% 
        collect()
        
head(wea_df)
```


## Get all data for specified station and year range
```{r}
the_sta <- 'KBOS'
year1 <- 1950
year2 <- 2016

wea_df <- tbl(wea_db,'wea_clean') %>% 
        filter(st_code==the_sta & between(year,year1,year2) & !is.na(yday)) %>% 
        select(date,st_code,year,yday,mean_temp) %>% 
        collect()

wea_df$date <- as.Date(wea_df$date)
#head(wea_df)

wea_df %>%  ggplot(aes(date,mean_temp)) +
        geom_line() +
        geom_smooth(method='lm')

```


## Compute yearly average temps
```{r}

the_sta <- 'KBOS'
year1 <- 1950
year2 <- 2016

wea_yearly <- tbl(wea_db,'wea_clean') %>% 
        filter(st_code==the_sta & between(year,year1,year2) & !is.na(yday)) %>% 
        distinct() %>% 
        select(date,st_code,year,yday,mean_temp) %>% 
        group_by(year) %>% 
        summarise(mt = mean(mean_temp)) %>% 
        collect()


wea_yearly %>%  ggplot(aes(year,mt)) +
        geom_point() +
        geom_smooth(method='lm') +
        ggtitle(paste('Yearly average temperatures for ',the_sta)) +
        ylab('Temperature (F)')

```




```{r}
tbl(wea_db, sql("SELECT count(distinct(st_code)) as N_stations  FROM wea_clean"))
```


# Trends in yearly-avgerage temperatures
- Compute yearly average temperatures
- * Need to make sure we have good data for entire years?
- fit linear regression and find trend (slope)


## First, see how many years have complete or almost complete data
```{r}
the_sta <- 'KBOS'
year1 <- 1950
year2 <- 2016

wea_yearly <- tbl(wea_db,'wea_clean') %>% 
        filter(st_code==the_sta & between(year,year1,year2) & !is.na(yday)) %>% 
        distinct() %>% 
        select(date,st_code,year,yday,mean_temp) %>% 
        group_by(year) %>% 
        summarise(mt = mean(mean_temp), N=count()) %>% 
        collect()
#        filter(N>300) %>% 
#        collect() 

wea_yearly

wea_yearly %>% ggplot(aes(year,N)) +
        geom_point() +
        ggtitle('Number of good daily data points each year')

```




```{r}

the_sta <- 'KBOS'
year1 <- 1950
year2 <- 2016

wea_yearly <- tbl(wea_db,'wea_clean') %>% 
        filter(st_code==the_sta & between(year,year1,year2) & !is.na(yday)) %>% 
        distinct() %>% 
        select(date,st_code,year,yday,mean_temp) %>% 
        group_by(year) %>% 
        summarise(mt = mean(mean_temp), N=count()) %>% 
        filter(N>300) %>% 
        collect()

wea_yearly %>%  ggplot(aes(year,mt)) +
        geom_point() +
        geom_smooth(method='lm') +
        ggtitle(paste('Yearly average temperatures for ',the_sta)) +
        ylab('Temperature (F)')

mod <- lm(mt~year,data = wea_yearly)
summary(mod)

```




